# High-level summary of handling of dynamic indexing

Dynamic indexing involves cases where the indexing of a variable in the model involves an unknown and is not a constant at the time of model definition. E.g.,

```
code <- nimbleCode({
  for(i in 1:4) {
        y[i] ~ dnorm(mu[k[i]+1], 1)
        k[i] ~ dcat(p[1:4])
     }
     for(j in 2:5)
        mu[j] ~ dnorm(0, 1)
})

```

When we detect this scenario, we inject an additional hidden model variable/node, named along the lines of `.mu_unknownIndex`. Each `y[i]` is a dependent of `.mu_unknownIndex`, while each `mu[j]` has `.mu_unknownIndex` as a dependent. This hidden model variable allows us to easily account for the changing structure of the graph as the `k[i]` values change.

We handle calculations in an inefficient, but robust way. `getDependencies(mu[j])` returns all of the `y` values (because of the hidden node lying between 'mu' and 'y'), so any calculate and simulate calls will be carried out for all possible dependents, even if a given `y[i]` is not currently a dependent of the `mu[j]` that has been modified.

Calculate and simulate calls check that current dynamic index values are within the range of possible indexes for a given variable. 

Functionality for handling dynamic indexing is only turned on when `nimbleOptions(allowDynamicIndexing)` is TRUE in version 0.6-6.

# Overview of processing stages for dynamic indexing

Various stages of creating a model definition in `BUGS_modelDef.R` involve special processing for dynamic indexing. In some cases calls are made to code in `BUGS_BUGSdecl.R`.

 - `genSymbolicParentNodes`: We first detect the presence of a dynamic index in `getSymbolicParentNodesRecurse`, called from `genSymbolicParentNodes`. At this point, the parent nodes are of a form like this (for the case of the above example):
    - `.USED_IN_INDEX(k[i])`
    - `.mu_unknownIndex_mu_oBk_oBi_cB_plus_1_cB_context2[.DYN_INDEXED(k[i] + 
    1)]`
    Note that it appears that `.USED_IN_INDEX()` is only used for a check for nonscalar dynamic indexing. IIRC the original intention was that this would be used for dynamically updating the graph when such a node was modified.
    
 - `genUnknownIndexDeclarations`: This adds a temporary declaration to `declInfo` that declares the unknownIndex variable implicitly introduced in the previous step. Having this new declaration results in the unknownIndex 'variable' get created in the graph using our regular processing. Dynamically indexed variables are detected based on the `.DYN_INDEXED()` tagging. Plus, the declaration is needed to ensure the original indexed variable still appears for processing if it is a RHS only (e.g., if `mu` were RHS only in the example). The new declaration looks like:
    - `.mu_unknownIndex_mu_oBk_oBi_cB_plus_1_cB_context2[.DYN_INDEXED(k[i] + 1)] <- mu[.DYN_INDEXED(k[i] + 1)]`
    - the `type` of the declaration is `unknownIndex`
    - vertex splitting in `genExpandedNodeAndParentNames3` is done based on this new declaration
    
 - `genVarInfo3`: `varInfo` for unknownIndex variables is not handled here (see next bullet), but note that for dynamically-indexed RHS variable, we do not learn anything about the variable dimensions because we do not look at the possible values of the dynamic index. 
 - `addUnknownIndexVars`:
    - Create variables for unknownIndex variables in varInfo. This is done here and not in `genVarInfo3` because `genVarInfo3` does a bunch of processing we don't need for the unknownIndex vars. unknownIndex vars are found based on `type=='unknownIndex'`
    - Also, add a vector, `unknownIndexNames`, to the modelDef that stores the unknownIndex variable names
 - `findDynamicIndexParticipants`: based on the `.DYN_INDEXED()` tagging, find dynamically indexed variables and:
    - add `dynamicIndexInfo` field to declInfo, containing the code that calculates the dynamic index and the range of possible index values based on the dimension of the indexed variable; thisinfo will be used to generate checks on legitimate index values in nodeFunctions
    - in `symbolicParentNodes`, replace dynamic indexing in `[ ... ]` with `[NA]`
    - in `symbolicParentNodes`, strip out `.USED_IN_INDEX()`
    
 - `addFullDimExtentToUnknownIndexDeclarations`: based on `type=='unknownIndex'` put full range of variable into targetExpr and symbolicParentNodes, e.g., `mu[1:5]` and `.mu_unknownIndex_mu_oBk_oBi_cB_plus_1_cB_context2[1:5]`. This info can then be used in our standard graph processing so that dependencies going through the unknownIndex var are handled properly with our usual processing.
 - `genExpandedNodeAndParentNames3`: as a result of our standard graph processing, the dependents of a dynamically-indexed node become dependents of the entire unknownIndex variable and the unknownIndex variable becomes a dependent of all relevant nodes that are dynamically-indexed. unknownIndex nodes are treated as LHSinferred so they are graph dependency calcs pass through them. 
 - `stripUnknownIndexDeclarations`: remove the unknownIndex BUGS declarations added in `genUnknownIndexDeclarations` (based on `type=='unknownIndex'`) and the unknownIndex vars added to varInfo in `addUnknownIndexVars` (based on `unknownIndexNames`). 



# Some older notes:

## Some prototype cases:

```{r, eval = FALSE}
## Basic case
quote({
    for(i in 1:5) x[i] ~ dnorm(0, sd = 1)
    p <- rep(0.2, 5)
    for(i in 1:10) {
        k[i] ~ dcat(p)
        y[i] ~ dnorm( x[k[i]], sd = 1)
    }
})


```



```{r, eval=FALSE}
## Computed indices
quote({
    for(i in 1:5) x[i] ~ dnorm(0, sd = 1)
    p <- rep(0.2, 4)
    for(i in 1:10) {
        k[i] ~ dcat(p)
        y[i] ~ dnorm( x[k[i]+1], sd = 1) ## or foo(k[i])
    }
})
```


```{r, eval=FALSE}
## computed from multiple other nodes
quote({
    for(i in 1:5) x[i] ~ dnorm(0, sd = 1)
    p <- rep(0.2, 2)
    for(i in 1:10) {
        k[i] ~ dcat(p)
        j[i] ~ dcat(p)
        y[i] ~ dnorm( x[k[i]+j[i]], sd = 1) ## or foo(k[i], j[i])
    }
})
```


```{r, eval = FALSE}
## Non-simple alignment of indices
quote({
    for(i in 1:5) x[i] ~ dnorm(0, sd = 1)
    p <- rep(0.2, 5)
    for(i in 1:10) {
        k[i] ~ dcat(p)
        y[i] ~ dnorm( x[k[11-i]], sd = 1)
    }
})
```


```{r, eval = FALSE}
## chained cases
quote({
    for(i in 1:5) x[i] ~ dnorm(0, sd = 1)
    p <- rep(0.2, 5)
    for(i in 1:10) {
        k[i] ~ dcat(p)
        y[i] ~ dnorm( x[k[i]], sd = 1)
    }
    
    for(i in 1:10) {
        j[i] ~ dcat(p)
        z[i] ~ dnorm( y[j[i]], sd = 1)
    }
    
})

```


```{r, eval = FALSE}
## More complicated connected cases
quote({
    for(i in 1:5) x[i] ~ dnorm(0, sd = 1)
    p <- rep(0.2, 5)
    for(i in 1:10) {
        k[i] ~ dcat(p)
        y[i] ~ dnorm( x[k[i]], sd = 1)
    }

    for(i in 1:5) w[i] ~ dnorm(0, sd = 1)
    for(i in 1:10) {
        l[i] ~ dcat(p)
        m[i] <- foo(l[i], x[floor(i/2)])
        z[i] ~ dnorm( w[m[i]] + y[i], sd = 1)
    }
    
})

```

## Design ideas

Using the first simple prototype as an example: There would be a special abstract node `x` representing all of x.  Let's call it a "dynamically indexed variable".  If helpful, we could label it `x_dynamicIndex_` or similar.   In effect all nodes within x (`x[i]`) would have x as a child.  And all nodes that use `x` with unknown index (`y[i]`) would have x as a parent.  But we need a way to get children of `x[3]` without including all of y.  So this would be done at run-time.

Note that we are already very close to having the capability of getting dependencies at run-time.  So this would be an extension of that.

In order to do so, we could have a system where, when we need dependencies of `c(x[3], x[5:7])` (i.e., an arbitrary subset of x), the occurrence of a dynamicIndex variable would trigger different handling.  First, in setup code, `model$getDependencies(c(x[3], x[5:7]))` would return an object that simply contains the content `c(x[3], x[5:7])` and a tag (attribute or class) that the dependencies need run-time evaluation.  Second, the run-time evaluation system would need to handle `x` specially: it would need to query its children for which ones currently use an index in `(3, 5, 6, 7)`.

I think the easiest way to implement this would be to lift non-simple dynamic indices.  E.g.,
```{r, eval = FALSE}
## from
y[i] ~ dnorm( x[ k[i] + 1], 1)
## we would create
y_index1_x[i] <- k[i] + 1
y[i] ~ dnorm( x[ y_index1_x[i] ], 1)
```
The point is that we need a way at run-time to ask for the index value, e.g. `k[i] + 1`.  Another way to do this would be to pack it into the `getParam` system, since this contains a `switch` statement with arbitrary computations for different cases.

Of course, the `getDependencies` input could start "above" elements of x and could continue beyond elements of y.   The "more complicated connected cases" is particularly challenging: A change to an element of x can have `y[i]` as a child, `m[i]` as a child (an index for `z[i]`), and then `z[i]` as a child by virtue of modifying `y[i]` and/or the index `m[i]`.  Note that I am using `i` generically here as "unkown index".

At first glance, this would imply that some nodes (e.g. `m[i]`) need to be calculated before subsequent dependencies can be determined.  In other words, determining dependencies and running the calculations would need to be intermixed.  For example, if we use the "current" value of `m[i]` to determine which `z[i]`s to include, would we get the wrong answer because the value of `m[i]` could change when the calculations are run?  But the following observation makes it possible to determine dependencies prior to any calculations: `z[5]` needs to count as a dependency *either* if `m[5]` is a dependency *or*  if `w[ current value of m[5] ]` is a dependency.   Put another way: say `w[3]` changes and `m[5] = 3`.   If `m[5]` changes, then `z[5]` must be a dependency because its index for `y` changed.  If `m[5]` does not change, then `z[5]` must be a dependency because `w[m[5]]` changed.

A topic for more thought is multivariate dynamic indexing and combinations of dynamic and static indexing.  From `x[i, k[j] ]`, would we create `x_dynamicIndex_i1_`, `x_dynamicIndex_i2_`, etc?  It may be feasible since these only exist as abstractions in the graph; we don't actually build nodeFunctions for them.  Also it would prevent wasted dynamic dependency searches over both indices (if we just decide any dynamic index makes all indices treated as dynamic).

## Implementation steps

This is going to be rough and surely not anticipate every step.

1. Modify the extraction of parentNodeNames to recurse into index expressions.  Currently that step errors out if an index involves a node.

2. Modify the extraction of parentNodeNames to somehow annotate when a parent involves a dynamic index (and in which index slot(s) this occurs). 

3. Modify construction of the graph information in R to include dynamicIndex nodes.  This would involve a new node type label and some additional thought: what information needs to be retained, if any?  We already keep track of which piece of a child expression each parent expression represents.  That may be enough.

4. Modify C++ graph classes.

5. Modify getDependencies so during setup it can return something annotated with "do this at run-time"

6. Modify C++ nodeFunctionVector to capture "do this at run-time"

7. Write versions of calculate etc. that gather dependencies at run time.  *A major part of this step can be done fairly independently of the previous steps.*   This would also allow us to benchmark the run-time cost of getting dependencies by doing so for static cases.


